---
title: "ATAC-seq Analysis"
author: "Bruna Toledo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

# Overview

This notebook documents the ATAC-seq analysis workflow used in the study.

It is designed as a lightweight execution notebook that calls the modular scripts stored in:

`ATAC_seq/scripts/`

The full workflow includes:

1. removal of unwanted variation from consensus peak counts  
2. differential accessibility analysis and genomic annotation  
3. PCA visualization  
4. gene-level accessibility heatmap  
5. GREAT enrichment analysis  
6. motif enrichment analysis  
7. motif enrichment plotting for TFs that are also RNA-seq DEGs  
8. ATAC–RNA integration analysis  

# Recommended repository structure

# Important note

This notebook assumes that the project files are organized exactly as expected by the scripts.
Each section below executes one script using `source()`.

# 1. Remove unwanted variation

This step imports consensus peak counts, filters low-count peaks, performs normalization, and estimates unwanted variation factors with RUVSeq.

```{r step-01}
source("../scripts/01_remove_unwanted_variation.R")
```

# 2. Differential accessibility and annotation

This step identifies differential accessible regions (DARs), annotates peaks, and summarizes genomic region classes.

```{r step-02}
source("../scripts/02_differential_accessibility_and_annotation.R")
```

# 3. PCA visualization

This step generates PCA plots from normalized ATAC-seq accessibility counts.

```{r step-03}
source("../scripts/03_pca_visualization.R")
```

# 4. Heatmap of accessibility

This step aggregates normalized accessibility to gene level and generates a heatmap with genomic region annotations.

```{r step-04}
source("../scripts/04_heatmap_accessibility.R")
```

# 5. GREAT enrichment

This step performs GREAT enrichment analysis on differentially accessible regions.

```{r step-05}
source("../scripts/05_great_enrichment.R")
```

# 6. Motif enrichment

This step evaluates TF motif enrichment across DARs stratified by genomic region.

```{r step-06}
source("../scripts/06_motif_enrichment.R")
```

# 7. Motif enrichment plot restricted to RNA-seq TF DEGs

This step visualizes motif enrichment restricted to transcription factors that are also differentially expressed in RNA-seq.

```{r step-07}
source("../scripts/07_motif_enrichment_deg_tfs_plot.R")
```

# 8. ATAC–RNA integration

This step integrates DARs with RNA-seq DEGs and performs GREAT enrichment by integration category.

```{r step-08}
source("../scripts/08_atac_rna_correlation.R")
```

# Session information

```{r session-info}
sessionInfo()
```
