---
title: "16S Microbiota Analysis"
author: "Bruna Toledo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

# Overview

This notebook illustrates the main 16S rRNA analyses performed in the study:

**“Short-term antibiotic treatment alters human blood neutrophil activation and gene regulation through microbiota disruption.”**

This document is intended as an analysis notebook for visualization and inspection of key results.  
The fully reproducible workflow is implemented in the scripts located in:

`16S_rRNA/scripts/`

Recommended execution order:

- `01_build_phyloseq.R`
- `02_diversity.R`
- `03_taxonomic_composition.R`
- `04_differential_abundance.R`
- `05_functional_prediction.R`

## Important note

This notebook does **not** rebuild the full workflow from raw data.  
It loads intermediate objects generated by the scripts above.

To knit this notebook successfully, first run the scripts in the order listed above.

# Load libraries

```{r load-libraries}
suppressPackageStartupMessages({
  library(MicrobiotaProcess)
  library(phyloseq)
  library(ggplot2)
  library(dplyr)
  library(tibble)
})
```

# Define input files

```{r define-inputs}
mpse_div_file <- "../results/intermediate/mpse_diversity.rds"
mpse_tax_file <- "../results/intermediate/mpse_taxonomy.rds"

aldex_results_file <- "../results/intermediate/aldex2_results.rds"
aldex_sig_results_file <- "../results/intermediate/aldex2_significant_results.rds"

picrust2_metacyc_sig_file <- "../results/intermediate/PICRUSt2_MetaCyc_ALDEx2_significant.rds"
picrust2_kegg_sig_file <- "../results/intermediate/PICRUSt2_KEGG_ALDEx2_significant.rds"
```

# Check required files

```{r check-files}
required_files <- c(
  mpse_div_file,
  mpse_tax_file,
  aldex_results_file,
  aldex_sig_results_file,
  picrust2_metacyc_sig_file,
  picrust2_kegg_sig_file
)

missing_files <- required_files[!file.exists(required_files)]

if (length(missing_files) > 0) {
  stop(
    paste(
      "The following intermediate files were not found:",
      paste(missing_files, collapse = "\n"),
      "",
      "Please run the scripts in `16S_rRNA/scripts/` before knitting this notebook.",
      sep = "\n"
    )
  )
}
```

# Load processed objects and result tables

```{r load-data}
mpse_div <- readRDS(mpse_div_file)
mpse_tax <- readRDS(mpse_tax_file)

aldex_results <- readRDS(aldex_results_file)
aldex_sig_results <- readRDS(aldex_sig_results_file)

picrust2_metacyc_sig <- readRDS(picrust2_metacyc_sig_file)
picrust2_kegg_sig <- readRDS(picrust2_kegg_sig_file)
```

# Rarefaction

Rarefaction was used to standardize sequencing depth across samples prior to diversity analyses.

```{r rarefaction-plot}
p_rarefaction <- mpse_div %>%
  mp_plot_rarecurve(
    .rare = RareAbundanceRarecurve,
    .alpha = Observe,
    .group = condition,
    plot.group = TRUE
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Pre-Abx" = "#3C548866", "Post-Abx" = "#87255B66")) +
  scale_color_manual(values = c("Pre-Abx" = "#3C548866", "Post-Abx" = "#87255B66"))

p_rarefaction
```

# Alpha diversity

Alpha diversity was calculated from rarefied abundance data to assess within-sample microbial diversity before and after antibiotic treatment.

```{r alpha-diversity}
p_alpha <- mpse_div %>%
  mp_plot_alpha(
    .group = condition,
    .alpha = c(Observe, Shannon, Pielou),
    test = "wilcox.test",
    test.args = list(paired = TRUE),
    textsize = 3.5,
    comparisons = list(c("Pre-Abx", "Post-Abx"))
  ) +
  scale_fill_manual(values = c("#3C548866", "#87255B66"), guide = "none") +
  scale_color_manual(values = c("#3C548866", "#87255B66"), guide = "none")

p_alpha
```

# Beta diversity

Beta diversity was assessed using Bray–Curtis dissimilarity and visualized by principal coordinate analysis (PCoA).

```{r beta-diversity}
p_beta <- mpse_div %>%
  mp_plot_ord(
    .ord = pcoa,
    .group = condition,
    .color = condition,
    .size = 4.8,
    .alpha = 0.75,
    ellipse = TRUE,
    show.legend = FALSE,
    show.adonis = TRUE
  ) +
  scale_fill_manual(values = c("#3C548866", "#87255B66")) +
  scale_color_manual(values = c("#3C548866", "#87255B66"))

p_beta
```

# Taxonomic composition

Relative abundance plots were used to summarize major shifts in microbial community composition across samples and experimental groups.

```{r plot-labels}
plot_label <- c(
  "S2_T1"  = "S2",  "S3_T1"  = "S3",  "S4_T1"  = "S4",  "S5_T1"  = "S5",
  "S6_T1"  = "S6",  "S7_T1"  = "S7",  "S8_T1"  = "S8",  "S9_T1"  = "S9",
  "S10_T1" = "S10", "S11_T1" = "S11", "S12_T1" = "S12", "S13_T1" = "S13",
  "S15_T1" = "S15", "S16_T1" = "S16",
  "S2_T2"  = "S2",  "S3_T2"  = "S3",  "S4_T2"  = "S4",  "S5_T2"  = "S5",
  "S6_T2"  = "S6",  "S7_T2"  = "S7",  "S8_T2"  = "S8",  "S9_T2"  = "S9",
  "S10_T2" = "S10", "S11_T2" = "S11", "S12_T2" = "S12", "S13_T2" = "S13",
  "S15_T2" = "S15", "S16_T2" = "S16"
)
```

## Phylum level

```{r phylum-composition}
p_phylum <- mpse_tax %>%
  mp_plot_abundance(
    .abundance = RareAbundance,
    .group = condition,
    taxa.class = Phylum,
    topn = 7,
    relative = TRUE,
    order.by.feature = FALSE
  ) +
  scale_x_discrete(labels = plot_label)

p_phylum
```

## Genus level

```{r genus-composition}
p_genus <- mpse_tax %>%
  mp_plot_abundance(
    .abundance = RareAbundance,
    .group = condition,
    taxa.class = Genus,
    topn = 20,
    relative = TRUE,
    order.by.feature = FALSE
  ) +
  scale_x_discrete(labels = plot_label)

p_genus
```

## Group-level summary

```{r phylum-group-summary}
p_phylum_group <- mpse_tax %>%
  mp_plot_abundance(
    .abundance = RareAbundance,
    .group = condition,
    taxa.class = Phylum,
    topn = 7,
    plot.group = TRUE
  )

p_phylum_group
```

# Differential abundance

Differential abundance analysis was performed with ALDEx2 across multiple taxonomic ranks.

## Significant genera

```{r aldex-genus-table}
if ("Genus" %in% names(aldex_sig_results)) {
  head(aldex_sig_results$Genus)
}
```

## Significant families

```{r aldex-family-table}
if ("Family" %in% names(aldex_sig_results)) {
  head(aldex_sig_results$Family)
}
```

## Example effect-size plot

```{r aldex-genus-plot}
if ("Genus" %in% names(aldex_sig_results) && nrow(aldex_sig_results$Genus) > 0) {

  genus_df <- aldex_sig_results$Genus %>%
    mutate(
      direction = ifelse(effect > 0, "Post-Abx", "Pre-Abx"),
      direction = factor(direction, levels = c("Pre-Abx", "Post-Abx"))
    )

  p_aldex_genus <- ggplot(genus_df, aes(x = reorder(feature, effect), y = effect, fill = direction)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Differential abundance at genus level",
      x = "",
      y = "Effect size",
      fill = ""
    ) +
    scale_fill_manual(values = c("Pre-Abx" = "#3C548866", "Post-Abx" = "#87255B99")) +
    theme_minimal() +
    theme(legend.position = "bottom")

  p_aldex_genus
}
```

# Functional prediction

PICRUSt2 was used to infer predicted microbial functional profiles from 16S rRNA data.  
Differential pathway abundance was then assessed with ALDEx2.

## MetaCyc pathways

```{r metacyc-results}
head(picrust2_metacyc_sig)
```

```{r metacyc-plot}
if (nrow(picrust2_metacyc_sig) > 0) {

  metacyc_plot_df <- picrust2_metacyc_sig %>%
    filter(!is.na(description)) %>%
    slice_max(order_by = abs(effect), n = min(20, nrow(.))) %>%
    mutate(
      condition_enriched = factor(condition_enriched, levels = c("Pre-Abx", "Post-Abx")),
      description = reorder(description, effect)
    )

  p_metacyc <- ggplot(metacyc_plot_df, aes(x = description, y = effect, fill = condition_enriched)) +
    geom_col(color = "black", width = 0.6) +
    coord_flip() +
    scale_fill_manual(values = c("Pre-Abx" = "#3C548866", "Post-Abx" = "#87255B66")) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.y = element_text(size = 10),
      legend.position = "bottom",
      legend.title = element_blank()
    ) +
    labs(
      title = "Differentially abundant MetaCyc pathways",
      x = "MetaCyc pathway",
      y = "Effect size"
    )

  p_metacyc
}
```

## KEGG pathways

```{r kegg-results}
head(picrust2_kegg_sig)
```

```{r kegg-plot}
if (nrow(picrust2_kegg_sig) > 0) {

  kegg_plot_df <- picrust2_kegg_sig %>%
    filter(!is.na(pathway_name)) %>%
    slice_max(order_by = abs(effect), n = min(20, nrow(.))) %>%
    mutate(
      condition_enriched = factor(condition_enriched, levels = c("Pre-Abx", "Post-Abx")),
      pathway_name = reorder(pathway_name, effect)
    )

  p_kegg <- ggplot(kegg_plot_df, aes(x = pathway_name, y = effect, fill = condition_enriched)) +
    geom_col(color = "black", width = 0.6) +
    coord_flip() +
    scale_fill_manual(values = c("Pre-Abx" = "#3C548866", "Post-Abx" = "#87255B66")) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.y = element_text(size = 10),
      legend.position = "bottom",
      legend.title = element_blank()
    ) +
    labs(
      title = "Differentially abundant KEGG pathways",
      x = "KEGG pathway",
      y = "Effect size"
    )

  p_kegg
}
```

# Notes on reproducibility

This notebook illustrates the main analyses and reproduces representative outputs from the 16S rRNA workflow.

The fully reproducible analysis pipeline is implemented in:

`16S_rRNA/scripts/`

Raw sequencing data are provided separately. Intermediate objects are not included in the repository and must be generated by running the scripts before knitting this notebook.

# Session information

```{r session-info}
sessionInfo()
```
